Uncertainty propagation in deep neural networks for sparse coding

### 2018-03-09 paper/BayesianLISTA.tex
soft-thresholding -> soft thresholding

\author{ {\bf Danil Kuzin} \\
\And
{\bf Olga Isupova}  \\
\And
{\bf Lyudmila Mihaylova}   \\
}

@@

\begin{abstract}
In the Bayesian neural network framework prior distributions are imposed on weights of a network and a posterior distribution of an output is inferred, rather than a point prediction provided by conventional neural networks. Therefore, Bayesian neural networks estimate both an output for a test data and uncertainty of this prediction. Uncertainty of the prediction is a result of propagating weight uncertainty via layers of a network. We propose a novel method of propagating uncertainty through the soft-thresholding nonlinearity. The method expands the area of Bayesian neural networks and enables the soft-thresholding nonlinearity to be included in Bayesian deep neural networks for the first time. As an example we design a Bayesian Learned Iterative Shrinkage-Thresdholding Algorithm (Bayesian \textsc{lista}) network. We develop a highly efficient posterior inference algorithm based on probabilistic backpropagation. A series of experiments on sparse coding show that the proposed framework provide both accurate predictions and quantified estimates of uncertainty in these predictions.
\end{abstract}

@@

Neural networks address the sparse coding problem with the soft-thresholding nonlinearity, that shifts current estimates towards zero \citep{gregor2010learning}. In this paper we propose a method to propagate uncertainty through the soft-thresholding nonlinearity. At every layer the current distribution of the target vector is represented as a spike-and-slab distribution \citep{mitchell1988bayesian}. We show that it can be effectively combined with Gaussian weights of dense layers. The distribution of the target vector propagated through a dense layer and soft-thresholding nonlinearity can then be closely approximated with a distribution from the same family.

The proposed method of uncertainty propagation for the soft-thresholding nonlinearity allows us to derive the gradients of the logarithms of normalisation constants, which can be used to update weights distribution. We then develop a highly efficient Bayesian inference algorithm based on the probabilistic backpropagation paradigm while significantly expanding it.

The main contributions of this paper can be summarised~as: 
\begin{itemize}
\item for the first time we propose a method for uncertainty propagation through the soft-thresholding nonlinearity for a Bayesian neural network
\item we develop efficient posterior inference for weights and outputs of neural networks with the soft-thresholding nonlinearity
\item we design a novel Bayesian neural network for sparse coding
\end{itemize}

@@

Lemma \ref{thm:matrix_const} describes the probabilistic variant of \textsc{lista} step $\quad \mathbf{b}_l = \mathbf{W}_l \mathbf{y}$, lemma \ref{thm:matrix_vector} and \ref{thm:sum_vectors} describe the probabilistic variant of \textsc{lista} step $ \mathbf{c}_l = \mathbf{b}_l +\mathbf{S}$ and lemma \ref{thm:soft_thresholding} describes the soft-thresholding step $\widehat{\boldsymbol\beta}_{l} = h_\lambda(\mathbf{c}_l)$. Overall the probabilistic layer is described by theorem \ref{thm:prob_layer}.

@@

\begin{lemma}[Gaussian propagation through soft thresholding]
\label{thm:soft_thresholding}
When a Gaussian-distributed random variable $x \sim \mathcal{N}(x; m, v)$ is propagated through the soft-thresholding function it can be approximated with the spike and slab distribution on $x^*$ with the probability of spike $\omega^*$, slab mean $m^*$ and slab variance $v^*$ (parameters of the approximating distribution are given in the proof).
\end{lemma}
\begin{proof}

The probability of spike $\omega^*$ equals to the probability mass of the original distribution that lies in $(-\lambda, \lambda)$  and is flattened into zero by soft-thresholding operator. As the original distribution is Gaussian, this can be computed as 
\begin{align}
\begin{split}
\omega^* &= p(x^*=0) = p(x \in [-\lambda, \lambda]) \\
&= \Phi\left(\frac{\lambda-m}{\sqrt{v}}\right) - \Phi\left(\frac{-\lambda-m}{\sqrt{v}}\right). 
\end{split}
\end{align}

Soft-thresholding shifts elements that are greater than $\lambda$  or less than $-\lambda$ elements towards 0. Let $\psi(\cdot)$ denote the density of the soft-threshold distribution, $\phi(\cdot)$ denote the density of the original Gaussian distribution. Then the first moment of the resulting distribution is

@@

\subsection{Approximation quality}
\label{sec:approx_quality}
We have used two approximations in forward propagation of uncertainty. First, in lemma \ref{thm:matrix_vector} a Gaussian matrix is multiplied by a spike-and-slab vector and their product is approximated with the Gaussian distribution. Second, in lemma \ref{thm:soft_thresholding} the result of soft-thresholding of a Gaussian vector is approximated with the spike-and-slab distribution. In this section we demonstrate that these approximations are close to the real distributions.

>>>>

%\author{ {\bf Danil Kuzin} \\
%\And
%{\bf Olga Isupova}  \\
%\And
%{\bf Lyudmila Mihaylova}   \\
%}

@@

\begin{abstract}
In the Bayesian neural network framework prior distributions are imposed on weights of a network and a posterior distribution of an output is inferred, rather than a point prediction provided by conventional neural networks. Therefore, Bayesian neural networks estimate both an output for a test data and uncertainty of this prediction. Uncertainty of the prediction is a result of propagating weight uncertainty through layers of a network. We propose a novel method of propagating uncertainty through the soft thresholding nonlinearity. The method expands the area of Bayesian neural networks and enables the soft thresholding nonlinearity to be included in Bayesian deep neural networks for the first time. As an example we design a Bayesian Learned Iterative Shrinkage-Thresdholding Algorithm (Bayesian \textsc{lista}) network. We develop a highly efficient posterior inference algorithm based on probabilistic backpropagation. A series of experiments on sparse coding shows that the proposed framework provide both accurate predictions and quantified estimates of uncertainty in these predictions.
\end{abstract}

@@

Neural networks address the sparse coding problem with the soft thresholding nonlinearity, that shifts current estimates towards zero \citep{gregor2010learning}. In this paper we propose a method to propagate uncertainty through the soft thresholding nonlinearity. At every layer the current distribution of the target vector is represented as a spike-and-slab distribution \citep{mitchell1988bayesian}. We show that it can be effectively combined with Gaussian weights of dense layers. The distribution of the target vector propagated through a dense layer and soft thresholding nonlinearity can then be closely approximated with a distribution from the same family.

The proposed method of uncertainty propagation for the soft thresholding nonlinearity allows us to derive the gradients of the logarithms of normalisation constants, which can be used to update weights distribution. We then develop a highly efficient Bayesian inference algorithm based on the probabilistic backpropagation paradigm while significantly expanding it.

The main contributions of this paper can be summarised~as: 
\begin{itemize}
\item for the first time we propose a method for uncertainty propagation through the soft thresholding nonlinearity for a Bayesian neural network;
\item we develop efficient posterior inference for weights and outputs of neural networks with the soft thresholding nonlinearity;
\item we design a novel Bayesian neural network for sparse coding.
\end{itemize}

@@

Lemma \ref{thm:matrix_const} describes the probabilistic variant of \textsc{lista} step $\quad \mathbf{b}_l = \mathbf{W}_l \mathbf{y}$, lemma \ref{thm:matrix_vector} and \ref{thm:sum_vectors} describe the probabilistic variant of \textsc{lista} step $ \mathbf{c}_l = \mathbf{b}_l +\mathbf{S}$ and lemma \ref{thm:soft_thresholding} describes the soft thresholding step $\widehat{\boldsymbol\beta}_{l} = h_\lambda(\mathbf{c}_l)$. Overall the probabilistic layer is described by theorem \ref{thm:prob_layer}.

@@

\begin{lemma}[Gaussian propagation through soft thresholding]
\label{thm:soft_thresholding}
When a Gaussian-distributed random variable $x \sim \mathcal{N}(x; m, v)$ is propagated through the soft thresholding function it can be approximated with the spike and slab distribution on $x^*$ with the probability of spike $\omega^*$, slab mean $m^*$ and slab variance $v^*$ (parameters of the approximating distribution are given in the proof).
\end{lemma}
\begin{proof}

The probability of spike $\omega^*$ equals to the probability mass of the original distribution that lies in $(-\lambda, \lambda)$  and is flattened into zero by soft thresholding operator. As the original distribution is Gaussian, this can be computed as 
\begin{align}
\begin{split}
\omega^* &= p(x^*=0) = p(x \in [-\lambda, \lambda]) \\
&= \Phi\left(\frac{\lambda-m}{\sqrt{v}}\right) - \Phi\left(\frac{-\lambda-m}{\sqrt{v}}\right). 
\end{split}
\end{align}

Soft thresholding shifts elements that are greater than $\lambda$  or less than $-\lambda$ elements towards 0. Let $\psi(\cdot)$ denote the density of the soft thresholded distribution, $\phi(\cdot)$ denote the density of the original Gaussian distribution. Then the first moment of the resulting distribution is

@@

\subsection{Approximation quality}
\label{sec:approx_quality}
We have used two approximations in forward propagation of uncertainty. First, in lemma \ref{thm:matrix_vector} a Gaussian matrix is multiplied by a spike-and-slab vector and their product is approximated with the Gaussian distribution. Second, in lemma \ref{thm:soft_thresholding} the result of soft thresholding of a Gaussian vector is approximated with the spike-and-slab distribution. In this section we demonstrate that these approximations are close to the real distributions.
